---
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{setspace}\doublespacing
indent: true
---

Jack Schroeder

Ec 1152  

Prof. Raj Chetty  

TF: Mike Droste  

Due 21 February 2019  

\begin{center}
Empirical Project 1
\end{center}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The main library I work with is the tidyverse.
# There are some other libraries I use (specifically when looking at covariates),
# but those libraries are called when they are used.

library(tidyverse)

# The atlas_ranks csv file is read in as oppatlas. I filter out NA
# values in each of the kfr pooled values. These NA values were not
# in New Jersey, so it should not impact the deeper analysis.

oppatlas <- read_csv("Files/atlas_ranks.csv") %>% 
  filter(!is.na(kfr_pooled_p25)) %>% 
  filter(!is.na(kfr_pooled_p75)) %>% 
  filter(!is.na(kfr_pooled_p100))

# I create a state database by filtering oppatlas for New Jersey.

state <- oppatlas %>% 
  filter(state == 34)

# I also create a Union County database, saved as county.

county <- state %>% 
  filter(county == 039)

# My hometown, Summit, can be found as a subset of Union County. 
# Summit's tracts are 37700, 37800, 37900, and 38000.

town <- county %>% 
  filter(tract == 037700 | tract == 037800 | 
           tract == 037900 | tract == 038000)

# Finally, I filter county to find my Census tract.

tract <- county %>% 
  filter(tract == 037700)

# One last thing in this setup panel. The following line turns off scientific notation,
# which makes R show the actual number when displaying output.

options(scipen=999)

```
My hometown, Summit, is located in Union County, New Jersey. It has a reputation of being a fairly well-off town with strong links to Wall Street, as 20% of its residents  work in finance and real estate (Patterson 2008). Its train station is a part of NJ Transit's Midtown Direct service, which operates express trains to Penn Station (Carter 2018). The Regional Plan Association found in 2010 that this service boosted property values in Summit by $23,000 for houses within walking distance of the train station (Carter 2018). Summit has a top-25 public school system in the state and is located close to many secular/religious private schools (U.S. News 2018). Summit's downtown suffered a downtown in the recession (Patterson 2008), but as of early 2019, its retail vacancy rate had decreased to 1.3% (Radest 2019). On the surface, Summit is resurgent and thriving.
```{r question 1, echo=FALSE}

# The outcomes I analyzed were household income and incarceration ranks.
# I saved each map from the Opportunity Atlas as a png file.
# I need the png library to read in png files from the Opportunity Atlas.

library(png)

# From there I just need to read in each png file and assign it.

q1graph1 <- readPNG("Files/opp-income-all.png")
q1graph2 <- readPNG("Files/opp-income-low.png")
q1graph3 <- readPNG("Files/opp-incarc-all.png")
q1graph4 <- readPNG("Files/opp-incarc-low.png")
```

In many outcomes, Summit is ahead of the rest of Union County. Of the 108 Census tracts within the county, Summit's tracts rank 2nd, 31st, 45th, and 53rd in the pooled household incomes of low-income children (`kfr_pooled_p25`). When displayed cartographically, Summit's outcomes for low-income children appear better than those of many other towns in the county. My home tract is highlighted in black, while the rest of Union County is southeast of that tract.

-----
Figure 1: Household Income among Low-Income Children in North Jersey
------

```{r figure 1, echo=FALSE, fig.align="center"}

# This if statement allows me to rasterize the image and display it
# in markdown. The fig.align code centers the photo.

if (exists("rasterImage")) {
plot.new()
rasterImage(q1graph2, 0,0,1,1)
}
```

When looked at through another metric, such as incarceration rates among low-income children, Summit still appears ahead of the pack. Within Union County, Municipalities like Cranford and Elizabeth have considerably worse outcomes than Summit. Stretching outward, the area around Newark, while located outside Union County, has strikingly worse outcomes in both household income and incarceration rates for low-income children.

-----
Figure 2: Incarceration Rates among Low-Income Children in North Jersey
------

```{r figure 2, echo=FALSE, fig.align="center"}

# I rasterize and show the second image in markdown with the following if statement.

if (exists("rasterImage")) {
plot.new()
rasterImage(q1graph4, 0,0,1,1)
}
```

```{r question 2, echo=FALSE}
# What period do the data
# you are analyzing come from? Are you concerned that the neighborhoods you are
# studying may have changed for kids now growing up there? What evidence do Chetty et
# al. (2018) provide suggesting that such changes are or are not important? What type of
# data could you use to test whether your neighborhood has changed in recent years?

# SDMTools provides the necessary functions to find population-weighted averages.

library(SDMTools)

# The code below calculates the weighted poverty rate in Summit in 1990, 2000, and 2010.
# An assumption made here is that population numbers have remained constant, which is not
# true. However, any form of population-weighing is better than the alternative of taking
# the mean of the four tracts (which have different populations).

wtmean <-town %>% select(tract, poor_share2010, poor_share2000, poor_share1990,
                          count_pooled)
wtmean1 <- wt.mean(wtmean$poor_share1990, wtmean$count_pooled)
wtmean2 <- wt.mean(wtmean$poor_share2000, wtmean$count_pooled)
wtmean3 <- wt.mean(wtmean$poor_share2010, wtmean$count_pooled)

# To demonstrate that the population-weighing did not skew the results, I also calculate
# the unweighted means. They still show an increase in poverty in Summit.

unweightedmean1 <- mean(town$poor_share1990)
unweightedmean2 <- mean(town$poor_share2000)
unweightedmean3 <- mean(town$poor_share2010)
```

Chetty et al. (2018)'s Opportunity Atlas data looks at children born between 1978 and 1983 and provides insights on their outcomes through adulthood. Using this data to look at Summit's outcomes may be misleading, since Summit itself has changed since the 1980s. After a population boom in the 1950s, the population began to precipitously decline in the 1970s. This decline continued until the late 1990s, but the city's population has yet to reach its high of 23,677 in 1960 (Census Data). Many in Summit believe that the city's population recovery was accompanied by fundamental changes in the town's identity, completing a transition into an upper-class Wall Street exclave in Union County. <ARE THESE CHANGES IMPORTANT> There are a few covariates that the Opportunity Atlas tracks that could help determine how much Summit has changed. For instance, the shares of each race are kept from both the 2000 and 2010 Census. The poverty rates from 1990, 2000, and 2010 are also in the Atlas' databases. From this data, we can see that Summit's poverty rate increased from 2% in 1990 to 4% in 2000, and then to 6% in 2010.

```{r question 3, echo=FALSE}
# Now turn to the atlas.dta data set. How does average upward mobility, pooling races and
# genders, for children with parents at the 25th percentile (kfr pooled_p25) in your
# home Census tract compare to mean (population-weighted, using count_pooled)
# upward mobility in your state and in the U.S. overall? Do kids where you grew up have
# better or worse chances of climbing the income ladder than the average child in America?

# I find weighted means for the nation, state, county, and home tract using wt.mean.
# I then round the results to help construct the table below.
# These averages look at children born in the 25th percentile.

q3.1 <- wt.mean(oppatlas$kfr_pooled_p25, oppatlas$count_pooled)
q3.1 <- round(q3.1, digits=0)
q3.2 <- wt.mean(state$kfr_pooled_p25, state$count_pooled)
q3.2 <- round(q3.2, digits=0)
q3.3 <- wt.mean(county$kfr_pooled_p25, county$count_pooled)
q3.3 <- round(q3.3, digits=0)
q3.4 <- wt.mean(tract$kfr_pooled_p25, tract$count_pooled)
q3.4 <- round(q3.4, digits=0)
```

```{r question 4, echo=FALSE}

# I then find standard deviations for nation, state, county, and home tract using wt.sd.
# I also round these results for the table.
# This is still only looking at children born in the 25th percentile.

q4.1 <- wt.sd(oppatlas$kfr_pooled_p25, oppatlas$count_pooled)
q4.1 <- round(q4.1, digits=0)
q4.2 <- wt.sd(state$kfr_pooled_p25, state$count_pooled)
q4.2 <- round(q4.2, digits=0)
q4.3 <- wt.sd(county$kfr_pooled_p25, county$count_pooled)
q4.3 <- round(q4.3, digits=0)

# There is no standard deviation for my home tract, so I find the standard deviation
# of my town to put in the table. This process is repeated below.

q4.4 <- wt.sd(town$kfr_pooled_p25, town$count_pooled)
q4.4 <- round(q4.4, digits=0)
```

```{r question 5, echo=FALSE}

# I find means and standard deviations in household income for children born in the 75th
# and 100th percentiles. I weigh these using wt.mean and wt.sd. They are then rounded
# to help with the table below.

q5.1 <- wt.mean(oppatlas$kfr_pooled_p75, oppatlas$count_pooled)
q5.1 <- round(q5.1, digits=0)
q5.2 <- wt.mean(state$kfr_pooled_p75, state$count_pooled)
q5.2 <- round(q5.2, digits=0)
q5.3 <- wt.mean(county$kfr_pooled_p75, county$count_pooled)
q5.3 <- round(q5.3, digits=0)
q5.4 <- wt.mean(tract$kfr_pooled_p75, tract$count_pooled)
q5.4 <- round(q5.4, digits=0)

q5.5 <- wt.sd(oppatlas$kfr_pooled_p75, oppatlas$count_pooled)
q5.5 <- round(q5.5, digits=0)
q5.6 <- wt.sd(state$kfr_pooled_p75, state$count_pooled)
q5.6 <- round(q5.6, digits=0)
q5.7 <- wt.sd(county$kfr_pooled_p75, county$count_pooled)
q5.7 <- round(q5.7, digits=0)
q5.8 <- wt.sd(town$kfr_pooled_p75, town$count_pooled)
q5.8 <- round(q5.8, digits=0)

q5.9 <- wt.mean(oppatlas$kfr_pooled_p100, oppatlas$count_pooled)
q5.9 <- round(q5.9, digits=0)
q5.10 <- wt.mean(state$kfr_pooled_p100, state$count_pooled)
q5.10 <- round(q5.10, digits=0)
q5.11 <- wt.mean(county$kfr_pooled_p100, county$count_pooled)
q5.11 <- round(q5.11, digits=0)
q5.12 <- wt.mean(tract$kfr_pooled_p100, tract$count_pooled)
q5.12 <- round(q5.12, digits=0)

q5.13 <- wt.sd(oppatlas$kfr_pooled_p100, oppatlas$count_pooled)
q5.13 <- round(q5.13, digits=0)
q5.14 <- wt.sd(state$kfr_pooled_p100, state$count_pooled)
q5.14 <- round(q5.14, digits=0)
q5.15 <- wt.sd(county$kfr_pooled_p100, county$count_pooled)
q5.15 <- round(q5.15, digits=0)
q5.16 <- wt.sd(town$kfr_pooled_p100, town$count_pooled)
q5.16 <- round(q5.16, digits=0)
```

Using the Opportunity Atlas, the average household income for children in the 25th, 75th, and 100th percentiles can be calculated using `kfr_pooled_p25`, `kfr_pooled_p75`, `kfr_pooled_p100`. These means, along with their standard deviations in parentheses, are displayed for my home tract, Union County, New Jersey, and the entire nation in Table 1. Since a single tract cannot have a standard deviation, I calculated the standard deviation of Summit's 4 tracts for the table.

Regardless of a child's percentile at birth, the average household income in my home tract is substantially higher than that of Union County, New Jersey, and the country. Union County itself is relatively average compared to the state totals (for each percentile rank), and New Jersey is slightly better-off than the rest of the country in each rank. The standard deviations in this table are also of interest because they denote mobility. My home tract has above-average mobility for children born in the 25th percentile, but that mobility falls as we look at higher percentiles. When looking at the 100th percentile, there is remarkably low mobility compared to Union County, New Jersey, and the country. This finding implies that children born into very wealthy families in my tract remained wealthy throughout life.

-----
Table 1: Outcome Matrix with Means and Standard Deviations
-----

| Percentile  | Tract | County | State | Nation |
| ----  | -----| ------| ------| ------|
| 25th  | \$`r q3.4` (\$`r q4.4`) | \$`r q3.3` (\$`r q4.3`) |  \$`r q3.2` (\$`r q4.2`) | \$`r q3.1` (\$`r q4.1`) |
| 75th  | \$`r q5.4` (\$`r q5.8`) | \$`r q5.3` (\$`r q5.7`) | \$`r q5.2` (\$`r q5.6`) | \$`r q5.1` (\$`r q5.5`) |
| 100th | \$`r q5.12` (\$`r q5.16`) | \$`r q5.11` (\$`r q5.15`) | \$`r q5.10` (\$`r q5.14`) | \$`r q5.9` (\$`r q5.13`) |


```{r question 6, echo=FALSE}
# Using a linear regression, estimate the relationship between outcomes of children at the
# 25th and 75th percentile for the Census tracts in your home county. Generate a scatter
# plot to visualize this regression. Do areas where children from low-income families do
# well generally have better outcomes for those from high-income families, too?

# This code runs a linear regression on child outcomes from the 25th and 75th percentiles.
# The data looks at Union County.

q6regression <- lm(kfr_pooled_p25 ~ kfr_pooled_p75, data=county)

# The regression yields a line of:
# kfr_pooled_p25 = 958.6194 (intercept) + 0.6876 * kfr_pooled_p75.

# The chunk below creates a scatter plot of the result. I use ranks to help with
# graph interpretation (I personally think it looks better). I multiply each rank
# to avoid having to manually set the axes. I also make each point lighter to help
# make the regression line (and its confidence) stand out more.

q6graph <- ggplot(county, aes(ranks_pooled_p75*100, ranks_pooled_p25*100)) +
  geom_point(alpha=.25) +
  xlab("Percentile Rank of Children Born in 75th Percentile") +
  ylab("Percentile Rank of Children Born in 25th Percentile") +
  ggtitle("Figure 3: Upward Mobility in Union County, NJ", subtitle = "Percentile Rank Data from Opportunity Atlas") +
  theme_minimal() +
  geom_smooth(method="lm")

```


```{r figure 3, echo=FALSE, warning=FALSE}

# I call for the graph and show it as Figure 3.

q6graph
```


```{r question 7, echo=FALSE, warning=FALSE}
# Next, examine whether the patterns you have looked at above are similar by race. If there
# is not enough racial heterogeneity in the area of interest (i.e., data is missing for most
# racial groups), then choose a different area to examine.

# To show whether the patterns are similar by race, I use the original graph and add
# geom_smooths by each race. I retitle the graph as Figure 4 and add a caption to 
# explain each line.

q7graph <- q6graph + 
  geom_smooth(aes(ranks_asian_p75*100, ranks_asian_p25*100), method="lm", color="red", se=FALSE) + 
  geom_smooth(aes(ranks_black_p75*100, ranks_black_p25*100), method="lm", color="brown", se=FALSE) + 
  geom_smooth(aes(ranks_hisp_p75*100, ranks_hisp_p25*100), method="lm", color="dark green", se=FALSE) +
  geom_smooth(aes(ranks_white_p75*100, ranks_white_p25*100), method="lm", color="orange", se=FALSE) +
  ggtitle("Figure 4: Upward Mobility in Union County, NJ (By Race)") +
  scale_x_continuous(limits = c(35, 75)) + labs(caption="Blue line: County Average
                                                Red line: Asian Children
                                                Orange line: White Children
                                                Green line: Hispanic Children
                                                Brown line: Black Children")

```

```{r figure 4, echo=FALSE, warning=FALSE}

# I call for the graph as Figure 4.

q7graph
```

 
```{r question 8, echo=FALSE, include=FALSE}
# Using the Census tracts in your home county, can you identify any covariates which help
# explain some of the patterns you have identified above? Some examples of covariates
# you might examine include housing prices, income inequality, fraction of children with
# single parents, job density, etc. For 2 or 3 of these, report estimated correlation
# coefficients along with their 95% confidence intervals.

# install prerequisite packages
#install.packages(c("sandwich", "foreign", "lmtest","devtools"))
#devtools::install_github("jjchern/sysuse")

# library for robust (white) standard errors
library(foreign)
library(sandwich)
library(lmtest)

# run ols
covariate_model_1 <- lm(ranks_pooled_p25 ~ foreign_share2010, data= county)
summary(covariate_model_1)

# grab white (robust) ses from vcv matrix
se1 <- sqrt(diag(vcovHC(covariate_model_1, "HC1")))

# construct CIs white (robust) ses
alpha <- 0.05
tval1 <- qt(p=1-alpha/2, df=covariate_model_1$df.residual)
model_1_CI_ub <- covariate_model_1$coefficients + tval1 * se1
model_1_CI_lb <- covariate_model_1$coefficients - tval1 * se1

# display CIs for white (robust) ses
cbind(model_1_CI_lb, model_1_CI_ub)

cor(county$singleparent_share2000, county$ranks_pooled_p25)

# run ols
covariate_model_2 <- lm(ranks_pooled_p25 ~ singleparent_share2000, data= county)
summary(covariate_model_2)

# grab white (robust) ses from vcv matrix
se2 <- sqrt(diag(vcovHC(covariate_model_2, "HC1")))

# construct CIs white (robust) ses
alpha <- 0.05
tval2 <- qt(p=1-alpha/2, df=covariate_model_2$df.residual)
model_2_CI_ub <- covariate_model_2$coefficients + tval2 * se2
model_2_CI_lb <- covariate_model_2$coefficients - tval2 * se2

# display CIs for white (robust) ses
cbind(model_2_CI_lb, model_2_CI_ub)

# run ols
covariate_model_3 <- lm(ranks_pooled_p25 ~ singleparent_share2010, data= county)
summary(covariate_model_3)

# grab white (robust) ses from vcv matrix
se3 <- sqrt(diag(vcovHC(covariate_model_3, "HC1")))

# construct CIs white (robust) ses
alpha <- 0.05
tval3 <- qt(p=1-alpha/2, df=covariate_model_3$df.residual)
model_3_CI_ub <- covariate_model_3$coefficients + tval3 * se3
model_3_CI_lb <- covariate_model_3$coefficients - tval3 * se3

# display CIs for white (robust) ses
cbind(model_3_CI_lb, model_3_CI_ub)

```

```{r question 9, echo=FALSE}
# Open question: formulate a hypothesis for why you see the variation in upward mobility
# for children who grew up in the Census tracts near your home and provide correlational
# evidence testing that hypothesis.
# For this question, many covariates have been provided to you in the atlas.dta file, which
# are described under the “Characteristics of Census tracts” header in Table 1.
# You are welcome to use outside data that are not included in atlas.dta, but this is not
# required. Diane Sredl has created a research guide for our class that contains links to
# other data sources. You may wish to read this tutorial on how to add variables to a data
# set in Stata.

# This code allowed me to figure out the average tract's foreign-born
# population share across the country. The following code chunk did
# the same for Union County.

nationforeign <- oppatlas %>% 
  filter(!is.na(foreign_share2010))
nationforeign <- wt.mean(nationforeign$foreign_share2010, nationforeign$count_pooled)
nationforeign <- round(nationforeign, digits=2)

unionforeign <- wt.mean(county$foreign_share2010, county$count_pooled)
unionforeign <- round(unionforeign, digits=2)
```

Table 2 presents three covariates of ranked income among children born into the 25th percentile. I chose three covariates to analyze: `foreign_share2010`, `singleparent_share2000`, and `singleparent_share2010`.

-----
Table 2: 95% Confidence Intervals for Correlation Coefficients of Covariates
-----

| Covariate | Lower Bound | *Coefficient* | Upper Bound |
|------------------------------------|:----------:|:----------:|:----------:|
| Share of Population Born Outside the U.S. in 2010 | -0.304 | *-0.224* | -0.143 |
| Share of Single-Headed Households with Children in 2000 | -0.485 | *-0.419* | -0.354 |
| Share of Single-Headed Households with Children in 2010 | -0.303 | *-0.247* | -0.192 |

First, I calculated the correlation between the foreign-born population share in each Union County tract and percentile rank of children born into the 25th percentile (the coefficient from `ranks_pooled_p25 ~ foreign_share2010`). There is a significant negative coefficient of -0.224 (95% confidence intervals are detailed in Table 2). Union County has a substantially higher foreign-born population share than the average Census tract (`r unionforeign` versus `r nationforeign` nationwide). The negative correlation suggests that poorer immigrants in Union County do not see much mobility. Indeed, when looking at percentile ranks, it appears that their children are in many cases worse-off than their parents.

Second, I looked at the share of single-headed households with children in 2000 (the coefficient from `ranks_pooled_p25 ~ singleparent_share2000`). This significant negative relationship has a correlation coefficient of -0.419.

Third, I looked at the share of single-headed households with children in 2010 (the coefficient from `ranks_pooled_p25 ~ singleparent_share2010`). The relationship here is also signficant and negative, with a coefficient of -0.247, but the impact is smaller than from 2000. This makes sense logically: the children whose outcomes are being analyzed were born around 1980, so the share of single parents in 2010 should not have as much of an effect on outcomes as the share in 2000.



```{r question 10, echo=FALSE}
# Putting together all the analyses you did above, what have you learned about the
# determinants of economic opportunity where you grew up? Identify one or two key
# lessons or takeaways that you might discuss with a policymaker or journalist if asked
# about your hometown. Mention any important caveats to your conclusions; for example,
# can we conclude that the variable you identified as a key predictor in the question above
# has a causal effect (i.e., changing it would change upward mobility) based on that
# analysis? Why or why not?
```

